---
title: "JSC370 Midterm Project"
author: "Zhengdan Li"
date: "03/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

```{r load-data, include=FALSE}
library(data.table)
library(dplyr)
library(leaflet)
library(ggplot2)
library(gridExtra)
library(grid)
library(gtable)
library(mgcv)
library(kableExtra)
library("GGally")
library(AICcmodavg)

# Helper functions
# Slice from end of the string
slice_back <- function(s, n) {
  sub = substr(s, nchar(s)-n+1, nchar(s))
  return(sub)
}
# Mode
fmode <- function(x) unique(x)[which.max(table(x))]

# Position order constant
position_levels <- c("Opposite Spiker", "Outside Hitter", "Middle Blocker",
       "Setter", "Libero")

# Read in the player and team data
scorers <- data.table::fread("vnl_2019/best-scorers.csv")
spikers <- data.table::fread("vnl_2019/best-spikers.csv")
diggers <- data.table::fread("vnl_2019/best-diggers.csv")
setters <- data.table::fread("vnl_2019/best-setters.csv")
receivers <- data.table::fread("vnl_2019/best-receivers.csv")

blockers <- data.table::fread("vnl_2019/best-blockers.csv")

bio <- data.table::fread("vnl_2019/player_bio.csv")
team_rank <- data.table::fread("vnl_2019/team_rank.csv")
round_robin <- data.table::fread("vnl_2019/round_robin.csv")

# Check data type and dimension 
str(scorers)
str(bio)
str(team_rank)
str(round_robin)
summary(scorers)
summary(bio)
summary(team_rank)
summary(round_robin)
```

Volleyball is among the most popular team sports around the world. The players and the audience enjoy it so much because of the strength, sportsmanship, and team collaboration demonstrated in the game. However, one may wonder that, to what extent is volleyball a team sport? In other words, which between a player's capability and team cooperation contributes more to winning a game?

This question further breaks down to the following subquestions to be discussed in this report:\
- How much does the score a player wins in a game depend on her personal capacity, such as physical traits and experience?\
- Is there a particular position or "ace" in the team critical for winning the game?\
- Is there an association between the scores two team members at specific positions get in a game?\

In this project, I analyzed data from the 2019 FIVB Volleyball Women's Nations League (VNL) to better understand the mechanisms of volleyball. Although the most recent VNL competition was held in 2021, available information on the players is not comprehensive. I then selected the second last competition, which was 2019 VNL as the game was cancelled in 2020. To learn the characteristics of as many teams as possible, I decided to focus on the preliminary round, where each of the 16 teams played against every other team once in the round-robin stage. The player roster dataset contains the basic information of the 400 players from the 16 teams, such as the age, height, and the position they play in the team. The team rank dataset contains the rank of the teams in terms of how many matches they won. Lastly, the best player datasets contain the summary of the player's spike, block, set, and dig through the tournament, respectively. 

### Terminology

Before diving into the methods, some volleyball terms relevant to the analysis are listed as follows. [[Reference](https://www.volleyball.com/volleyball-101/volleyball-terms/)]\
- **Match**: Volleyball matches are a made up of sets. Match play on VNL consists of competing until one team wins 3 out of 5 sets.\
- **Set (game)**: A volleyball set is played to a 25 points. Sets must be won by at least 2 points.\
- **Dig**: Passing a rapidly hit ball from the rally (potentially to the setter).\
- **Set**: The tactical skill in which a ball is directed to a point where a player can spike it into the opponent's court.\
- **Spike**: The offensive action of hitting the ball. The attempt by one team to terminate the play by hitting the ball to the floor on the opponent's side. Also "attack".\
- **Middle Blocker**: Middle blockes hit balls set at the net in the middle of the court. Middle blockers are commonly used as a decoy to freeze or confuse the opposing teams blockers.\
- **Outside Hitter**: Outside hitters are usually the primary attackers on the team. These hitters attack balls that are set to the left side of the court.\
- **Setter**: Setters have the 2nd of 3 contacts of the ball and “set” the ball with an “Overhand Pass” for a attacker to hit.\
- **Libero**: liberos serves as defensive specialists on the team who usually "dig" with the 1st of 3 contacts of the ball. \
- **Opposite Spiker**: Opposite hitters, also know as right-side hitters, is considered the most versatile because they can excel on offense and defense.

## 2. Methods

### 2.1 Data Source
The data was retrieved by web scraping at [this link](https://en.volleyballworld.com/en/vnl/2019/). The Python library Beautiful Soup was used to pull the data from the web pages. The Python script and fetched datasets can be viewed on [this GitHub repository](https://github.com/Avery7Li/JSC370-project/tree/master).

### 2.2 Tools for Exploration
I read the datasets into data tables. The data cleaning and wrangling operations were completed using primarily data.table methods. Most figures preseted in this report were created using ggplot2. The GGally library was used to generate the pairwise correlation plots.

### 2.3 Data Cleaning
After the datasets are ready, I started exploratory data analysis. The team rank data table has 16 rows and 19 columns, which is expected as there are 16 teams in the tournament. The player bio data table contains 400 rows for 400 players and 13 columns, while the best scorer data table contains 245 rows. This is because many substitute players did not get a chance to play or score and were not included in the ranking. 

A further look into the variables revealed some issues with the original data table For the bio data table, three outliers of `spike height` lower than 240cm were observed from the simple boxplot. It turned out two of them are from liberos with height 162, which was reasonable because libero is the position that has the least requirement of height. However, the third outlier with value 130cm comes from a middle blocker of height 192cm, which is likely a mistake. Thus, I replaced the 130cm by 292cm, which is the block height from the same player. When I looked at the position variable from the bio data table, I saw that only one player has a position named "Universal" which is not among the positions introduced in the previous terminology. The player was from the DOM team and had a total score of 93. I then substituted her position by the mode position, outside hitter. In addition, the cases of the player position is not consistent. I standardized them to capital cases.

### 2.4 Data Wrangling

I noticed that some names on the bio data table had a letter "c" preceded by a newline at the end of the string. As I confirmed on the website, the "c" indicates that the player was the captain of the team. There were 16 such names as the competition had 16 team participants. I created a new indicator variable `is_captain` with value 0 and 1 before removing the suffix from the name string.

`age` was created as a new numerical variable. I extracted the year from the `birthdate` variable, converted it into an integer value, and subtracted it from year 2019 to calculate a player's age at the time of the competition.

Before merging the data tables together, I renamed the ambiguous columns with specific names. For example, variables called `total` were both present in the best scorer and bio datasets, but they represent total scorer and the number of times a player had been selected for competitions, respectively.

Next, I took variable `total score` from the best scorers dataset, `attack success rate` from the best spikers dataset, `digs per set` from the best  diggers dataset, `sets per set` from the best setters dataset, and `attack success rate` from the best attackers dataset. I merged the bio and with the variables of interest mentioned above using the player name. The combined data table still has 400 rows - Luckily, there were not two players with the same name. 

However, 112 rows have NA values in all columns `total score`, `digs per set`, `sets per set`, and `attack success rate`. Missing values were expected as previously we saw bio and best scorers have different row numbers. I first dropped the 112 rows and then imputed the remaining missing values with average within the group `position` because players from the same position are more likely to have similar scores. The result player data table has 288 rows. 

Then, to prepare for the later investigation of the association between player performance and team result, I merged the newly generated player data table with the team rank data table. This time, no missing values are introduced to the combined table.
 

```{r cleaning & wrangling, include=FALSE}
## bio
# Rename ambiguous columns before merging
setnames(bio, 'spike', 'spike_height')
setnames(bio, 'block', 'block_height')
setnames(bio, 'total', 'total_selections')

# Deal with outliers
boxplot(bio$spike_height, main='Boxplot of spike height')
bio[spike_height < 240]
nrow(bio[spike_height >= block_height]) / nrow(bio)
bio[, spike_height := fifelse(spike_height < 150, block_height, spike_height)]

# Only one player has a universal position, replace with the most occurred position
bio[position == 'Universal']
bio[, position := fifelse(position == 'Universal', fmode(position), position)]

# Make position name consistent
bio[, position := fifelse(position == 'Middle blocker', 
                          'Middle Blocker', 
                          fifelse(position == 'Opposite spiker',
                                  'Opposite Spiker', position))]

## Create numerical variables - bmi, world_game, age
bio[, bmi := round(weight / (height/100)**2, 1)]
bio[, world_selection := world_championships + olympic_games]
# Take the last four char from birth date and convert to int as birth year
bio[, birth_year := strtoi(slice_back(birthdate, 4))]
# Calculate age
bio[, age:= 2019 - birth_year]
# Create a indicator for captain: name has '\nc' at the end
bio[, is_captain := fifelse(slice_back(name, 2) == '\nc', 1, 0)]
# Remove char from end of name
bio[, name := gsub("\nc", "", name)]
# Keep only relevant columns

## scorers
# Rename columns
setnames(scorers, 'total', 'total_score')
setnames(scorers, 'rank', 'score_rank')
# Drop column that's not interesting
scorers <- scorers[, !"shirtnumber"]

# Merge player scores with bio
# look for association between players' individual scores and their ability
players <- merge(x = bio[, .(name, team, position, age, height, weight, bmi,
                             spike_height, block_height,
                             total_selections, is_captain)], 
                 y = scorers, 
                 all.x = T, all.y = F, 
                 by=c("name", 'team'))

## Best players
# Rename
setnames(diggers, 'average_per_set', 'digs_per_set')
setnames(setters, 'average_per_set', 'sets_per_set')
setnames(spikers, 'success_%', 'attack_success_rate')
# Merge with player dataset
players <- merge(x = players, 
                 y = diggers[, .(name, digs_per_set)],
                 all.x = T, all.y = F, 
                 by="name")
players <- merge(x = players, 
                 y = setters[, .(name, sets_per_set)], 
                 all.x = T, all.y = F, 
                 by="name")
players <- merge(x = players, 
                 y = spikers[, .(name, attack_success_rate)], 
                 all.x = T, all.y = F, 
                 by="name")
# Check row number doesn't change so no players have the same name
print(nrow(players) == nrow(bio))
# Check for NAs
players[is.na(digs_per_set) & is.na(sets_per_set) & 
         is.na(attack_success_rate) & is.na(total_score), .N]
players <- players[!(is.na(digs_per_set) & is.na(sets_per_set) & 
         is.na(attack_success_rate) & is.na(total_score))]
players[is.na(total_score), .N]   # 43

# Impute missing values
players <- players %>%
  group_by(position) %>%
  mutate(
    total_score = coalesce(total_score, round(mean(total_score, na.rm=TRUE))),
    digs_per_set = coalesce(digs_per_set, round(mean(digs_per_set, na.rm=TRUE))),
    sets_per_set = coalesce(sets_per_set, round(mean(sets_per_set, na.rm=TRUE)))
    #attack_success_rate = coalesce(attack_success_rate,
    #                               round(mean(attack_success_rate, na.rm=TRUE)))
  )
players <-data.table(players)

## Team
# Rename team rank
setnames(team_rank, 'rank', 'team_rank')
# Merge team info together with player info
team_players <- merge(x=team_rank[, .(team, team_rank, match_win, 
                                      set_ratio, team_full)], 
                      y=players[, .(team, age, position, height, 
                                    total_selections, total_score,
                                    digs_per_set, sets_per_set, 
                                    attack_success_rate)],
                      
                     all.x = T, all.y = T, 
                      by='team')

```


## 3. Preliminary Results

### 3.1 How much does the score a player wins in a game depend on her personal capacity, such as physical traits and experience?

Starting with the first question of interest, I focused on the variable `total score` as the measurement of performance, variables `height` and `spike height` for physical traits, and variables `age` and `total selection` (number of times the player had been selected to compete in championships) for the measurement of experience. The total score is composed of attack score, block score, and serve score. The correlation between spike height and block height is as large as 0.89. I also found that the spike height is no smaller than the block height for all players, so I picked `spike height` as an indicator for jump height. 

I created two sets of pairwise correlation plots of total score with physical measurements and experience measurements, respectively.

```{r individual correlation plot, echo=FALSE, width="80%"}
ggpairs(players[, .(total_score, height, spike_height)], 
        title="Pairwise Correlation Among Player Height, Spike Height, and Total Score",
        columnLabels = c("Total Score", "Height", "Spike Height"))
ggpairs(players[, .(total_score, total_selections, age)], 
        title="Pairwise Correlation Among Player Age, Number of Selections, and Total Score", 
        columnLabels = c("Total Score", "Number of Selections", "Age"))
```

We can observe on the plot that there is a large correlation between spike height and height of value 0.735. The correlation between total score and spike height is significant as indicated by the three asterisks, and the correlation coefficient is 0.403. In terms of experience, the number of selections is significantly associated with age. The total score the player won during the competition is positively correlated with the number of selections. The correlation has a coefficient value of 0.171, which is relatively weak compared to that with spike height. However, the performance of a player is not associated with age as the correlation value 0.003 is very close to zero.

Then, I fitted a linear model to total score and spike height. The estimated coefficient of spike height is 1.65, which represents that an increment of 1cm in spike height will result in an increase of 1.65 points in a player's score. The p-value is $1.10\times 10^{-12}$ smaller than the significance level $\alpha=0.05$, so the estimate is statistically significant. The adjusted $R^2$ of the multiple linear regression model is 0.16, so about 16% of the variation in total score can be explained by the independent variables spike height. 

However, I identified some problems from the diagnostic plots. The line on the residual vs fitted graph is a little curved rather than horizontally straight. Some points on the tails from the Normal QQ plot fall off the line. Thus, the residuals are not normally distributed. The linear model might not be an ideal fit for `total score` and `spike height`.

```{r individual lm, include=FALSE}
cor.test(players$spike_height, players$block_height) #0.89
cor.test(players$total_score, players$total_selections) #0.135

# Linear regression wrt height
# total selection is not significant
score_height_fit <- lm(total_score ~ spike_height, data=players)
summary(score_height_fit)
par(mfrow=c(2,2))
plot(score_height_fit)

# LM among attackers
attackers <- players[position == "Outside Hitter" | position == "Opposite Spiker" ]
score_height_fit_attacker <- lm(total_score ~ spike_height, data=attackers)
summary(score_height_fit_attacker)
par(mfrow=c(2,2))
plot(score_height_fit_attacker)

# Regression lines
# ignore
players %>%
  ggplot() +
  # Color points by position
  aes(x = spike_height, y = total_score) +
  geom_jitter() +
  # Fit a linear regression line by region
  stat_smooth(method=lm, formula = y~x) + 
  labs(title="Overall Regression Lines of Total Score vs Spike Height") + 
  theme_light() + 
  scale_color_brewer(palette = "Set2")
```

However, the score a player wins in a match depends on the nature of her position, and each position has different physical standard for the player. For example, opposite spikers and outside hitters are the primary attackers on the team and score most of the points. Those positions have high requirement on the player height as they need to hit the ball high to pass the block from the rally. On the other hand, liberos serve as defensive specialists and usually do not have an opportunity to score directly. They need to be agile to dig the ball before it hits the floor and thus have a lower weight among the players. Similarly, setters usually have the 2nd of the three contacts with the ball and pass the ball to attackers on the team. It is unusual setters score directly on the court except serves.

```{r score position summary, echo=FALSE, out.width="80%", fig.align="center"}
# Summary statistics, group by position
position_summary <- players[, .(
  `Average Score` = round(mean(total_score)),
  #`Average Attacks` = round(mean(attacks)),
  #`Average Blocks` = round(mean(blocks)),
  `Average Height` = round(mean(height)),
  `Average Weight` = round(mean(weight)),
  `Number` = .N
), by=position]

position_summary[order(-`Average Score`)] %>%
  kbl(caption="Summary Statistics of 5 Volleyball Positions") %>%
  kable_styling("striped", "hover", full_width = F)


# Violin plot of total score by position
players %>%
  ggplot() + 
    aes(y = total_score, x = 1, fill = position) +
    geom_violin() +
    facet_wrap(~position, nrow=1) +
    scale_fill_brewer(palette = "Set2") +
    labs(title="Distribution of Player Score Based on Positions",
         x="", y="Total Score") +
    theme_light() +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
          legend.position='none')

```

I visualized the total scores versus spike height by five volleyball positions as follows. It is noticeable that the regression lines are the steepest among opposite spikers and outside hitters. 

```{r individual plot, echo=FALSE, width="80%"}
# Note: Strongest correlation between total score and spike height 
# among opposite spikers and outside hitters
transform(players,
          position=factor(position, levels=position_levels))%>%
  ggplot() +
  # Color points by position
  aes(x = spike_height, y = total_score, color=position, alpha=0.5) +
  geom_jitter() +
  # Fit a linear regression line by region
  stat_smooth(method=lm, formula = y~x) + 
  labs(title="Scatterplots of Total Score vs Spike Height by Position",
       x="Spike Height", y="Total Score") + 
  facet_wrap(~position) +
  guides(color = 'none', alpha = 'none') +
  theme_light() + 
  scale_color_brewer(palette = "Set2")
```

Therefore, There is a moderate association between the points a player wins on a competition and her physical traits, especially with the jump height. Attackers on the team are likely to score points on a match. The association is stronger among some volleyball positions and weaker among others. However, a player's performance is not necessarily related to her experience. Young players who have not participated in a lot of world championships can gain as many points as experienced players.

### 3.2 Is there a particular position or "ace" in the team critical for winning the game?

To answer this question, I studied the association between the teams' results and players' performance. The team result was evaluated using the metric `match points` and `set ratio`. Match points were calculated by adding up points gained from each match according the following criteria.

```{r point criteria, echo=FALSE}
criteria <- data.table(case=c("Match won 3-0; 3-1", "Match won 3-2", 
                              "Match lost 2-3", "Match lost 0-3; 1-3", 
                              "Match forfeited"),
                       points=c("3 points ", "2 points ", "1 points ",
                                "0 points ", "0 points "))
criteria %>%
  kbl(caption="Points Earned Per Match") %>%
  kable_styling("striped", full_width = F)
```


The teams aimed to finish a match by winning three out of five sets as soon as possible ideally without losing a set. If two or more teams are tied on the number of points gained, they will be ranked by `set ratio` resulting from the division of the number of all set won by the number of all sets lost. Players' performance was measured based on their total score, success rate of attacks, the number of digs per set, and the number of sets per set. Actions of digs and sets are usually the first two contacts with the ball and not counted towards scores, but they also reflect player skills, especially for liberos and setters.

First, I took a closer look at each of the 16 teams. Summary statistics of match results, player performance, and play bio are shown in the table below. Maximum and average total scores of players in the team are calculated for each team together with the standard deviation. For digs and sets, because one of the 6 players on the court in a game will be a setter and at most one of them will be a libero, I only took maximum values on the number of digs and sets per set for each team.

```{r team table, echo=FALSE, out.width="80%", fig.align="center"}
# Team player summary statistics
team_player_summary <- players[, .(
  max_score = max(total_score),
  average_score = round(mean(total_score)),
  std_score = round(sd(total_score), 2),
  max_digs = max(digs_per_set),
  max_sets = max(sets_per_set),
  average_age = round(mean(age)),
  average_height = round(mean(height)),
  average_total_selections = round(mean(total_selections))
), by=team]
team_summary <- merge(x=team_rank[, .(team, team_rank, match_win, points,
                                      set_ratio, team_full)],
                      y=team_player_summary, by='team')
# Knitr table
team_summary[, .(team_rank, team_full, match_win, points, set_ratio, max_score, 
                 average_score, std_score, max_digs, max_sets, 
                 average_age, average_total_selections, average_height)][order(team_rank)] %>%
  kbl(col.names = c("Rank", "Team", "Match Win", "Match Points", "Set Ratio",
                    "Max Score", "Average Score", "STD Score", 
                    "Max Digs", "Max Sets", "Average Age", 
                    "Average Selections", "Average Height")) %>%
  column_spec(7, color = "white", 
              background = spec_color(team_summary$std_score, 
                                      option="D", end=0.7,
                                      alpha ="0.8", direction=1)) %>%
  kable_styling("striped", "hover") %>%
  add_header_above(c(" " = 2, "Match Result" = 3, 
                     "Player Performance" = 5, "Player Bio" = 3)) %>%
  scroll_box(height='500px')
```

In terms of player profile, Team Belgium was the youngest team with an average age of 22 while the largest average age was 29 in Team Thailand. At the same time, Thailand players were the shortest with an average height of 175cm. The largest average height of 188cm came from Team USA. While Team China and Team USA both won 12 matches, the former had a larger set ratio, which means Team China had lost fewer sets during the matches. However, neither Team China nor USA had a particularly large max or average player score. Team China's max player score was among the last five teams and they had the third smallest standard deviation of player scores. On the other hand, the best scorer of the season, Malwina Smarzek, won 421 points on her own for Team Poland, but the team ranked in 6th place, which was not as good as one may think. Therefore, the score an ace wins for the team is not deterministic. 

The spreads of players' scores for each team are shown in the following boxplots. Starting from the top left subplot, the teams are ordered according to the rank from left to right. 

```{r team plot, echo=FALSE, out.width="80%", fig.align="center", fig.align="center"}
# Boxplot of player total score by team
team_in_order <- team_rank[, team_full]
transform(team_players,
          team_full=factor(team_full,levels=team_in_order)) %>%
  ggplot() + 
    aes(y = total_score, x = 1) +
    geom_boxplot() +
    coord_flip() +
    facet_wrap(~team_full, nrow=4) +
    labs(title="Boxplots of Player Score in 16 Teams",
         x="", y="Total Score") +
    theme_light() +
    theme(axis.text.y=element_blank(), axis.ticks.y=element_blank())

# Hist of total selection by top 5 teams
team_in_order <- team_rank[, team_full]
transform(team_players,
          team_full=factor(team_full,levels=team_in_order))[team_rank <= 5] %>% 
  ggplot(aes(x=total_selections, fill=team_full)) +
  geom_histogram(binwidth = 50, position = 'dodge2') +
  labs(title="Histogram of Selections in the Top 5 Teams", 
       x="Number of Selection", y="Count") +
  scale_fill_brewer(palette = "Set2", name = "Team") +
  theme_light() 
```

We can observe that many teams have one or two players who are the primary scorer on the team. However, Having players at similar levels does not necessarily put a team at a disadvantage. Though it is better for the players to be similarly strong rather than similarly weak.

#TODO: comment on the histogram

```{r team and individual correlation, include=FALSE, out.width="80%", fig.align="center", warning=FALSE}
team_players[team=="CHN", .(position, total_score)]
team_players[team=="RUS", .(position, total_score)]
team_players[team=="ITA", .(position, total_score)]

# Correlation
cor.test(team_summary$max_score, team_summary$set_ratio)
cor.test(team_summary$average_score, team_summary$set_ratio)
cor.test(team_summary$std_score, team_summary$set_ratio)
#cor.test(team_summary$max_digs, team_summary$set_ratio) #negative
#cor.test(team_summary$max_sets, team_summary$set_ratio) #negative

# Based on individual player: set ratio with metrics - not significant
win_rate_all_fit <- lm(set_ratio ~ total_score + sets_per_set + 
                         digs_per_set, data=team_players)
summary(win_rate_all_fit)

# Based on team summary
models <- rep(0, 4)
r2_list <- rep(0, 4)
res_var <- rep(0, 4)
p_values <- rep(0, 4)
win_all_fit <- lm(points ~ average_score + max_sets + max_digs, data=team_summary)
summary(win_all_fit)

win_score_set_fit <- lm(points ~ average_score + max_sets, data=team_summary)
summary(win_score_set_fit)

win_score_dig_fit <- lm(points ~ average_score + max_digs, data=team_summary)
summary(win_score_dig_fit)

win_score_fit <- lm(points ~ average_score, data=team_summary)
summary(win_score_fit)
par(mfrow=c(2,2))
plot(win_score_fit)
models <- list(win_all_fit, win_score_set_fit, win_score_dig_fit, win_score_fit)
```

Next, I tried to fit linear models to match points the 16 teams won based on different combinations of explanatory variables of average play score, maximum number of digs per set, and maximum number of sets per set. The resulted coefficients of determination (adjusted), residual variance, model p-values, and AICc are shown in the following table. 

```{r team model summary, echo=FALSE, out.width="80%", fig.align="center", warning=FALSE}
# Display in table table
for (i in seq(1, 4)) {
  r2_list[i] <- round(summary(models[[i]])$adj.r.squared, 3)
  res_var[i] <- round(summary(models[[i]])$sigma^2, 3)
  p_values[i] <- round(summary(models[[i]])$coefficients[2,4], 3)
}

model_names <- c("Average Score + Max Digs + Max Sets",
            "Average Score + Max Sets", 
            "Average Score + Max Digs", "Average Score")
aic <- aictab(models, modnames = model_names)
win_score_summary <- data.table(
  `Explanatory Variables` = model_names,
  `Coefficient of Determination` = r2_list,
  `Residual Variance` = res_var,
  `Model p-value` = p_values,
  `AICc` = aic$AICc
)

win_score_summary %>%
  kbl(caption="Linear Models on Teams' Match Points") %>%
  kable_styling("striped", "hover", full_width = F)
```

We can notice that the linear model with average score and max digs has the largest adjusted $R^2$, which indicates that around 29% of the variation in the match points can be explained by the average score and maximum number of digs per set. That model also has the smallest residual variance among the four and a p-value smaller than the 0.05 significance level. However, the model fitted with all three independence variables is favoured by the AICc model selection.

The graph below shows the visualization of the linear regression of match points with the average player score. The plot hints that the linear model may not be a good fit for the data, as many points fall outside the shaded 95% interval. The model could not predict games with the particularly high or low points well.

```{r team regression plot, echo=FALSE, out.width="80%", fig.align="center"}
# Regression line of set ratio with player average score
team_summary %>%
  ggplot() +
  # Color points by position
  aes(x = average_score, y = points) +
  geom_jitter() +
  # Fit a linear regression line by region
  stat_smooth(method=lm, formula = y~x) + 
  labs(title="Regression Lines of Match Points vs Average Player Score",
       x="Average Player Score", y="Match Points") + 
  theme_light() 
# Note: GAM has no difference
```

Then, I explored how many best players each team had, according to the spike, block, set, and dig rankings. The scatterplot below decomposes the top 10 of the 4 skill rankings by the teams and player positions. The x-axis corresponds to the rank from 1 to 10, so each vertical line would have four pins for the four skills. The teams on the y-axis are ordered by the preliminary round team rank from top to bottom.

```{r skill plot, echo=FALSE, out.width="100%", fig.align="center"}
# how many best players in each team - heatmap of rank
# Wrangle
rank_all <- rbind(spikers[, .(name, rank, skill = "Attack", value=attack_success_rate)],
                  setters[, .(name, rank, skill = "Set", value=sets_per_set)],
                  diggers[, .(name, rank, skill = "Dig", value=digs_per_set)],
                  #receivers[, .(name, rank, skill = "Receive", value=`efficiency_%`)],
                  blockers[, .(name, rank, skill = "Block", value=average_per_set)])
rank_all <- merge(x=rank_all, 
                  y=bio[, .(team, name, position, is_captain)], 
                  all.x=T, all.y=F, by='name')
rank_all <- merge(x=rank_all, 
                  y=team_rank[, .(team_rank=rank, team_full, team)], 
                  all.x=T, all.y=F, by='team')
# Scatterplot
transform(rank_all,
          team_full=factor(team_full, levels=rev(team_in_order)))[rank <= 10] %>%
  ggplot(aes(x=rank, y=team_full, shape=position, color=skill, size=3)) + # is_captain
  #ggplot(aes(x=rank, y=team_full, color=position, shape=skill, size=3)) +
  geom_point() +
  #facet_wrap(~team_full, nrow=16) +
  scale_x_continuous(breaks=seq(1,10))+
  theme_light() +
  guides(size="none") +
  #theme(axis.ticks.y=paste(rank_all$team_rank, rank_all$team_full))
  scale_color_brewer(palette = "Set2") +
  labs(title="Composition of Top 10 Players with 4 Skills By Teams", 
       x="Player Skill Rank", y="Team")
rank_all_summary <- rank_all

team_in_order_short <- team_rank[,team]
transform(rank_all,
          team=factor(team, levels=team_in_order_short))[rank <= 10] %>%
  ggplot(aes(x=team, fill=skill)) +
  geom_bar() +
  geom_hline(yintercept=2.5, color="grey", linetype="dashed") +
  scale_fill_brewer(palette = "Set2") +
  theme_light() +
  labs(title="Barchart of Top Players by Teams and Skills", 
       x="Team", y="Count")
```

There is no surprise that the top 10 players who excel at setting were setters, and opposite spikers and outside hitters were the best at attacks. It is interesting to note that the 10 best diggers could be liberos, outside hitters, and even setters. Although players need to actively attack to score, preventing the ball from hitting the floor at the first place is equally important. Thus, it never hurts to have more people good at digging on the team.

USA has the best spiker, but not captain

```{r rank composition}
# composition of positions in the top 10 best player rank - pie facet
rank_all[is_captain==1 & rank <= 10, .(name, team, position, rank, skill)]
```




```{r captain, echo=FALSE, out.width="80%", fig.align="center"}
# Filter captains in the 16 teams
captains <- players[is_captain == 1]
mean(captains$age)
mean(captains$total_selections)
mean(captains$total_score)
boxplot(captains$age)

# Pie chart
captain_position <- captains[, .(number = .N), by=position]
captain_position <- 
  captain_position %>% 
  arrange(desc(position)) %>%
  mutate(prop = number / sum(captain_position$number) *100) %>%
  mutate(ypos = cumsum(prop)- 0.5*prop )
captain_pie <- captain_position %>%
  ggplot(aes(x="", y=prop, fill=position)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  geom_text(aes(y = ypos, label = number), color = "white", size=4) +
  scale_fill_brewer(palette="Set2") +
  theme_void()
# Histogram of captains' scores
captain_hist <- captains %>% 
  ggplot(aes(x=total_score, fill=position)) +
  geom_histogram(binwidth = 50) +
  labs(title="Histogram of Captains' Total Score by Position", 
       x="Total Score", y="Count") +
  scale_fill_brewer(palette = "Set2") +
  theme_light() +
  guides(color='none') +
  theme(legend.position = 'none')
grid.arrange(captain_pie, captain_hist, ncol=2)

#TODO: Score rank for captain within each team
captains$score_rank

```


### 2.3 Is there an association between the scores two team members at specific positions get in a game?

```{r position interaction within team, echo=FALSE, out.width="80%", fig.align="center"}

```


## 4. Conclusion


